A High Efficiency MAC Protocol for WLANs: Providing Fairness in Dense Scenarios

### 2015-10-22 introProtocol.tex
Second draft complete

CSMA/ECA~\cite{barcelo2008lba} is a fully distributed and collision-free MAC for WLANs. It differs from CSMA/CA in that it uses a deterministic backoff, $B_{\text{d}}=CW_{\min}/2$ after successful transmissions, where $CW_{\min}$ is the minimum contention window of typical value $CW_{\min}=16$. By doing so, contenders that successfully transmitted on schedule $n$, will transmit without colliding with other successful nodes in future cycles.

@@

Referring to Figure~\ref{fig:BECA}, CSMA/ECA is able to achieve an aggregated throughput that goes beyond CSMA/CA up until the number of contenders ($N$) is greater than $B_{\text{d}}$ ($B_{\text{d}}=7$ in the case of the figure). Beyond this point, the network will have a mixed behavior relating to backoff mechanisms: some nodes will successfully transmit and use a deterministic backoff while others will collide due to the lack of empty slots and return to a random backoff. As more contenders join the network, CSMA/ECA performance will approximate to CSMA/CA's. \textcolor{red}{This effect is also observed in L-ZC and L-MAC, albeit at a larger number of nodes. This is becasue the virtual cicle used by these protocolos ($N$ for L-ZC and $C$ for L-MAC in Section~\ref{relatedWork}) in greater than B$_{d}$. Specifically, the virtual cycle is of size equal to $16$ (changing to $B_{\text{d}}=16$ would only delay CSMA/ECA's throughput degradation to $N=B_{\text{d}}$).}

@@

	Algorithm~\ref{alg:fullECA} shows an implementation of CSMA/ECA$_{\text{Hys+FS}}$, while an example of CSMA/ECA$_{\text{Hys+FS}}$ with four contenders is shown in  Figure~\ref{fig:ECA+Hyst}. In the figure the first outline indicates a collision between STA-3 and STA-4, which will provoke an increment on both station's backoff stage ($k\leftarrow k+1$). Once STA-4's random backoff expires, CSMA/ECA$_{\text{Hys+FS}}$ instructs the station to transmit $2^{k}$ packets, and then use a deterministic backoff, $B_{\text{d}}=CW(k)/2$. The same behavior is followed by STA 3.

	With Hysteresis and Fair Share, CSMA/ECA$_{\text{Hys+FS}}$ is able to achieve greater throughput than CSMA/CA and for many more contenders, as shown also in Figure~\ref{fig:BECA}. In the figure, the \emph{CSMA/ECA$_{\text{Hys+FS}}$} curve shows a greater throughput because collisions are eliminated and Fair Share allows nodes to send $2^{k}$ packets upon each transmission. This throughput increase is the result of aggregation via Fair Share, which also carries the effect of raising the average time between successful transmissions (see Section~\ref{timeBetweenSxTx}), which may affect delay-sensitive traffic, like gaming or live video/voice/tv streaming. \textcolor{red}{Channel errors or unsaturated traffic conditions will disrupt any collision-free schedule, generate collisions and push all stations's backoff stages to the maximum value very quickly (contributing to the delay increase). This issue is leveraged with Schedule Reset mechanims, presented in Section~\ref{schedReset}.}

@@

	\subsection{Clock drift issue in descentralized collision-free MAC protocols}\label{clockDrift-issue}
	CSMA/ECA relies on stations being able to correctly count empty slots and consequently attempt transmissions in the appropriate slot according to the backoff timer. Failure to do so may be caused by clock imperfections inside the Wireless Network Interface Cards (WNIC), which is commonly referred to as \emph{clock drift}. As pointed out in~\cite{slotDrift}, clock drift is a common issue that degrades the throughput in distributed collision-free MAC protocols like the ones reviewed in Sect.~\ref{relatedWork}.

	While miscounting empty slots have no significant effect on CSMA/CA's throughput~\cite{slotDrift}, it has a direct impact on CSMA/ECA. In a collision-free schedule with saturated CSMA/ECA contenders, a station miscounting empty slots will \emph{drift} to a possibly busy slot, collide and force a re-convergence (if possible) to a collision-free schedule (see Sect.~\ref{performanceClockDrift}).

@@

	Failed transmissions due to channel errors also means that a few moments of operation under a noisy channel can increase the contenders's deterministic backoff to its maximum value. Such event carries the undesired effects of increasing the time between successful transmissions, and a reduction in the overall system throughput due to shorter collision-free periods.

@@

	Nevertheless, having a big deterministic backoff increases the time between successful transmissions. Furthermore, if not operating in a scenario with many nodes the empty slots between transmissions are not longer negligible and degrade the overall throughput of the system. For instance, if nodes withdraw from the contention their previously used slots will now be empty. On the other hand, in scenarios with channel errors contenders rapidly end up with the largest deterministic backoff. Nodes should be aware of this issue and pursue opportunities to reduce their deterministic backoff, $B_{\text{d}}$ without sacrificing too much in collisions.

@@

		\subsubsection{Keeping collision-free schedules}
		Schedule Reset is described in Algorithm~\ref{alg:schedRest}. Each contender starts filling a bitmap $b$ of size $B_{\text{d}}$ with the information observed in each passing slot during $\gamma=B_{\text{d}}/C$ consecutive successful transmissions (sxTx in Algorithm~\ref{alg:schedRest}); where $C=2^{m}CW_{\min}/2$ is the biggest deterministic backoff, and $\gamma$ is referred to as the Schedule Reset \emph{threshold} from this point forward.

@@

		Using Figure~\ref{fig:scheduleReset1} as an example, we can see that the sum $\sum\limits_{t} b[t]=0$, for $t\in\{8,16,24\}$. This means that the transmission slots corresponding to a deterministic backoff, $B^{*}_{\text{d}}=8$ are free, therefore a change of schedule is possible. In case of suffering a collision immediately after applying the schedule reduction, the node reverses the changes made by Schedule Reset before handling the collision.

@@

		\subsubsection{Facing channel errors}\label{aggr}

@@

		This effect is leveraged by setting $\gamma=1$, meaning that the bitmap is filled with the information of slots between only two consecutive transmissions. This measure increases the frequency of schedule reset attempts.
		%Further, there is the option of doing a schedule \emph{halving}. That is, with a set $A\in\{2^{k-1}CW_{\min}/2\}$.
	%	
	%	Schedule Reset can also be configured to increase the node's \emph{stickiness} after a successful reduction of the schedule. Stickiness is not a new concept to CSMA/ECA~\cite{barcelo2011tcf}. It simply instructs the contender to \emph{stick} to the deterministic backoff even in the event of \emph{stickiness} number of consecutive failed transmissions. This allows for a faster convergence towards a collision-free schedule, especially when performing under heavy channel errors. 
	%

>>>>

CSMA/ECA~\cite{barcelo2008lba} is a fully distributed and collision-free MAC for WLANs. It differs from CSMA/CA in that it uses a deterministic backoff, $B_{\text{d}}=CW_{\min}/2$ after successful transmissions, where $CW_{\min}$ is the minimum contention window of typical value $CW_{\min}=16$. By doing so, contenders that successfully transmitted on schedule $n$, will transmit without colliding with other successful nodes in future cycles~\cite{HE}.

@@

Referring to Figure~\ref{fig:BECA}, CSMA/ECA is able to achieve an aggregated throughput that goes beyond CSMA/CA up until the number of contenders ($N$) is greater than $B_{\text{d}}$ ($B_{\text{d}}=8$ in the case of the figure). Beyond this point, the network will have a mixed behavior relating to backoff mechanisms: some nodes will successfully transmit and use a deterministic backoff while others will collide due to the lack of empty slots and return to a random backoff. As more contenders join the network, CSMA/ECA performance will approximate to CSMA/CA's. \textcolor{red}{This effect is also observed in L-ZC and L-MAC. It happens becasue the virtual cicle used by these protocolos ($M$ for L-ZC and $C$ for L-MAC in Section~\ref{relatedWork}) in lower than the number of contenders, $N$.}

@@

	Algorithm~\ref{alg:fullECA} describes CSMA/ECA$_{\text{Hys+FS}}$, while an example of CSMA/ECA$_{\text{Hys+FS}}$ with four contenders is shown in  Figure~\ref{fig:ECA+Hyst}. In the figure the first outline indicates a collision between STA-3 and STA-4, which will provoke an increment on both station's backoff stage ($k\leftarrow k+1$). Once STA-4's random backoff expires, CSMA/ECA$_{\text{Hys+FS}}$ instructs the station to transmit $2^{k}$ packets, and then use a deterministic backoff, $B_{\text{d}}=CW(k)/2$. The same behavior is followed by STA 3.

	With Hysteresis and Fair Share, CSMA/ECA$_{\text{Hys+FS}}$ is able to achieve greater throughput than CSMA/CA and for many more contenders, as shown also in Figure~\ref{fig:BECA}. In the figure, the \emph{CSMA/ECA$_{\text{Hys+FS}}$} curve shows a greater throughput because collisions are eliminated and Fair Share allows nodes to send $2^{k}$ packets upon each transmission. This throughput increase is the result of aggregation via Fair Share, which also carries the effect of raising the average time between successful transmissions (see Section~\ref{timeBetweenSxTx}), which may affect delay-sensitive traffic, like gaming or live video/voice/tv streaming.

	\textcolor{red}{Channel errors, hidden-nodes, or unsaturated traffic conditions will disrupt any collision-free schedule, generate collisions and push all stations's backoff stages to the maximum value very quickly (contributing to the delay increase). This issue is leveraged with the Schedule Reset mechanim, introduced in Section~\ref{schedReset}.}

@@

	\subsection{Clock drift issue in descentralized collision-free MAC protocols}\label{clockDrift-issue}
	CSMA/ECA relies on stations being able to correctly count empty slots and consequently attempt transmissions in the appropriate slot according to the backoff timer. Failure to do so may be caused by clock imperfections inside the Wireless Network Interface Cards (WNIC), which is commonly referred to as \emph{clock drift}. As pointed out in~\cite{slotDrift}, clock drift is a common issue that degrades the throughput in distributed collision-free MAC protocols like the ones reviewed in Section~\ref{relatedWork}.

	While miscounting empty slots have no significant effect on CSMA/CA's throughput~\cite{slotDrift}, it has a direct impact on CSMA/ECA. In a collision-free schedule with saturated CSMA/ECA contenders, a station miscounting empty slots will \emph{drift} to a possibly busy slot, collide and force a re-convergence (if possible) to a collision-free schedule (see Section~\ref{performanceClockDrift}).

@@

	Failed transmissions due to channel errors also means that a few moments of operation under a noisy channel can increase the contenders's deterministic backoff to its maximum value. Such event carries the undesired effects of increasing the time between successful transmissions and reducing the overall system throughput due to shorter collision-free periods.

	\textcolor{red}{\subsection{Hidden terminals}}\label{hidden}
	This type of nodes posses an incomplete view of the network. Therefore, they disrupt any collision-free operation in CSMA/ECA$_{\text{Hys+FS}}$ by attempting transmissions while the channel is busy.

	For CSMA/ECA$_{\text{Hys+FS}}$, as for other reservation MAC protocols for CSMA (see Section~\ref{relatedWork},~\cite{HE,L_MAC}), hidden terminals affects the performance as channel errors do. That is, increasing the number of collisions and reducing the periods of collision-free operation.

@@

	Nevertheless, having a big deterministic backoff increases the time between successful transmissions. Furthermore, if not operating in a scenario with many nodes, the empty slots between transmissions are not longer negligible and degrade the overall throughput of the system. For instance, if nodes withdraw from the contention their previously used slots will now be empty. On the other hand, in scenarios with channel errors contenders rapidly end up with the largest deterministic backoff. Nodes should be aware of this issue and pursue opportunities to reduce their deterministic backoff, $B_{\text{d}}$ without sacrificing too much in collisions.

@@

		\subsubsection{A conservative schedule reduction}
		Schedule Reset is described in Algorithm~\ref{alg:schedRest}. Each contender starts filling a bitmap $b$ of size $B_{\text{d}}$ with the information observed in each passing slot during $\gamma=B_{\text{d}}/C$ consecutive successful transmissions (sxTx in Algorithm~\ref{alg:schedRest}); where $C=2^{m}CW_{\min}/2$ is the biggest deterministic backoff, and $\gamma$ is referred to as the Schedule Reset \emph{threshold} from this point forward.

@@

		Using Figure~\ref{fig:scheduleReset1} as an example, we can see that $\sum\limits_{t} b[t]=0$, for $t\in\{8,16,24\}$. This means that the transmission slots corresponding to a deterministic backoff, $B^{*}_{\text{d}}=8$ are free, therefore a change of schedule is possible. In case of suffering a collision immediately after applying the schedule reduction, the node reverses the changes made by Schedule Reset before handling the collision.

@@

		\subsubsection{Aggressiveness to face channel errors}\label{aggr}

@@

		This effect is leveraged by setting $\gamma=1$, meaning that the bitmap is filled with the information of slots between only two consecutive transmissions. This measure increases the frequency of schedule reset attempts. Furthermore, incrementing the node's stickiness after a successful reduction of the schedule allows for a faster convergence towards a collision-free schedule, especially when performing under heavy channel errors. 
	%

### 2015-10-22 journal.tex
Added new figures, that combined accumulate less pages than before. Still coexistence, and SR to go
Second draft complete

\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{float}
\usepackage{color, colortbl}
\usepackage{xcolor}
\usepackage{array}
\usepackage{multirow}
\usepackage{footnote}
\usepackage{cite}
%The below is used to add notes to tables without disrupting the IEEEtran format
\usepackage{threeparttable}
%Multiple figures in a row
%\usepackage{caption}
%\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{mathtools}
%Algorithms
\usepackage[linesnumbered,vlined]{algorithm2e}

@@

%Put the references in a nice way so the empty space is not overly great.
\usepackage{flushend}

@@

\textcolor{red}{This work describes CSMA/ECA and its extensions. Additionally, it provides the first evaluation results of CSMA/ECA with non-saturated traffic, channel errors, and its performance when coexisting with CSMA/CA nodes. Furthermore, it describes the effects of imperfect clocks over CSMA/ECA and present a mechanism to leverage the impact of channel errors over collision-free schedules. Finally, experimental results on throughput and lost packets from a CSMA/ECA implementation using commercial hardware and open-source firmware are presented.}
\end{abstract}

@@

\begin{itemize}
	\item First results on the achievable throughput and delay of CSMA/ECA with Hysteresis and Fair Share under non-saturated traffic conditions for very large number of nodes.
	\item The impact of imperfect clocks on CSMA/ECA with Hysteresis and Fair Share's deterministic backoff and its consequences on the achieved performance.
	\textcolor{red}{\item The effect of channel errors over collision-free schedules and how to solve the issue of converging rapidly to the biggest cycle length with the Schedule Reset mechanism.}
	\item Formulation of the throughput bounds for CSMA/ECA with Hysteresis and Fair Share.
	\item Coexistence and backwards compatibility with CSMA/CA nodes under different traffic conditions.
	\item First implementation of CSMA/ECA with Hysteresis in real hardware and the experimental results in a crowded WLAN testbed.
\end{itemize}

%Results show that CSMA/ECA with Hysteresis and Fair Share is capable of accommodating many users in collision-free schedules. Further, the coexistence tests show that the average aggregated throughput of a mixed network is higher than that observed in CSMA/CA-only networks, as well as a high resilience to clock-drifts.
\textcolor{red}{Results, derived from a modified version of the COST~\cite{COST} simulator, show that CSMA/ECA with Hysteresis and Fair Share is capable of accommodating many users in collision-free schedules. Further, as tests in a mixed network with different proportions of CSMA/CA and CSMA/ECA nodes show, the aggregated throughput is higher than the observed in CSMA/CA-only networks without significant sacrifice from legacy stations.}

@@

\section{Conclusions}\label{conclusions}
CSMA/ECA$_{\text{Hys+FS}}$ is able to construct a collision-free schedule with many contenders. Taking advantage of this condition, the cumulative throughput experienced by CSMA/ECA$_{\text{Hys+FS}}$ nodes goes beyond the achievable by CSMA/CA for any number of nodes. All of these while preserving throughput fairness. Further, as non-saturated CSMA/ECA$_{\text{Hys+FS}}$ networks get crowded the cumulative throughput keeps increasing up to the saturation point, contrary to the throughput degradation seen in CSMA/CA networks.

>>>>

\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{float}
\usepackage{color, colortbl}
\usepackage{xcolor}
\usepackage{array}
\usepackage{multirow}
\usepackage{footnote}
\usepackage{cite}
\usepackage{lipsum}
%The below is used to add notes to tables without disrupting the IEEEtran format
\usepackage{threeparttable}
%Multiple figures in a row
%\usepackage{caption}
%\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{mathtools}
%Algorithms
\usepackage[linesnumbered,vlined]{algorithm2e}

@@

%Put the references in a nice way so the empty space is not overly great.
%\usepackage{flushend}

@@

\textcolor{red}{This work describes CSMA/ECA and its extensions. Additionally, it provides the first evaluation results of CSMA/ECA with non-saturated traffic, channel errors, and its performance when coexisting with CSMA/CA nodes. Furthermore, it describes the effects of imperfect clocks over CSMA/ECA and present a mechanism to leverage the impact of channel errors over collision-free schedules. Finally, experimental results on throughput and lost frames from a CSMA/ECA implementation using commercial hardware and open-source firmware are presented.}
\end{abstract}

@@

\begin{itemize}
	\item First results on the achievable throughput and delay of CSMA/ECA with Hysteresis and Fair Share under non-saturated traffic conditions for very large number of nodes.
	\item \textcolor{red}{The impact of imperfect clocks and channel errors on CSMA/ECA with Hysteresis and Fair Share's deterministic backoff and its consequences on the achieved performance.}
	\textcolor{red}{\item Introduce the Schedule Reset mechanism for reducing the schedule length in case of users withdrawing from the contention (non-saturated traffic), or when its value is increased due to channel errors.}
	\item Formulation of the throughput bounds for CSMA/ECA with Hysteresis and Fair Share.
	\item Coexistence and backwards compatibility with CSMA/CA nodes under different traffic conditions.
	\item First implementation of CSMA/ECA with Hysteresis in real hardware and the experimental results in a crowded WLAN testbed.
\end{itemize}

%Results show that CSMA/ECA with Hysteresis and Fair Share is capable of accommodating many users in collision-free schedules. Further, the coexistence tests show that the average aggregated throughput of a mixed network is higher than that observed in CSMA/CA-only networks, as well as a high resilience to clock-drifts.
\textcolor{red}{Results, derived from a modified version of the COST~\cite{COST} simulator, show that CSMA/ECA with Hysteresis and Fair Share is capable of accommodating many users in collision-free schedules. Further, as tests in a mixed network with different proportions of CSMA/CA and CSMA/ECA nodes show, the aggregated throughput is higher than the observed in CSMA/CA-only networks.}

@@

\section{Conclusions}\label{conclusions}
CSMA/ECA$_{\text{Hys+FS}}$ is able to construct a collision-free schedule with many contenders. Taking advantage of this condition, the cumulative throughput experienced by CSMA/ECA$_{\text{Hys+FS}}$ nodes goes beyond the achievable by CSMA/CA for any number of nodes. All of these while preserving throughput f	airness. Further, as non-saturated CSMA/ECA$_{\text{Hys+FS}}$ networks get crowded the cumulative throughput keeps increasing up to the saturation point, contrary to the throughput degradation seen in CSMA/CA networks.

### 2015-10-22 realImplementation-Gringoli.tex
Second draft complete

We confirmed the simulations results experimentally with a testbed of real prototypes. This phase was mandatory to check whether the behaviour of the system is still correct in the presence of unexpected phenomena, like excessive channel noise, or technical limitations of the hardware like imperfect node synchronisation; and to verify the implementation feasibility of the Schedule Reset mechanism with a real time horizon.

As CSMA/ECA$_{\text{Hys}}$ works with the standard 802.11 PHY, we do not need the flexibility offered by complex development kits based on FPGAs like the WARP boards~\cite{amiri2007warp}: any architecture that allows a customisation of the channel access delay would fit our requirements. For this reason we chose a widely available 802.11b/g WNIC from Broadcom. Like many other cards, its behaviour is ruled by a firmware code that controls the underlying hardware (including the radio, carrier sense and FIFOs) in real time. In particular we used the BRCM4318KBFG PCI card as it is supported by the OpenFWWF~\cite{OpenFWWF} open-source firmware, an alternative to the original code from the manufacturer that has been recently considered as development platform in several research projects~\cite{WMP,gringolitmc14,gringoliccr14,berger14,CF-MAC}.

Thanks to the extremely low price of this card (refurbished devices are widely available for less than $\$10$ each), it is possible to run experiments with very large testbeds. We used 25 PC-Engines Alix 2d2 nodes running Linux kernel 2.6.32 as stations and a more powerful computer with the same kernel as access point, all nodes being equipped with the aforementioned WNIC.

@@

To incorporate stickiness into the prototype we added a {\tt STATE} to the system that can be either {\tt STATE\_DETERMINISTIC} or {\tt STATE\_RANDOM}, and a {\tt STICKINESS} counter that we reset each time we have a success and decrement when failing. When the counter gets to zero we enter the random state. Upon a successful transmission we unconditionally enter the deterministic state and reset the stickiness counter to {\tt DEFAULT\_STICKINESS}.
%Setting this last parameter to zero lets the system operate without the stickiness feature.

For the Schedule Reset mechanism, we added a bitmap for monitoring the state of every single slot after a successful transmission. To index the current slot in the bitmap we used the hardware register {\tt SPR\_IFS\_BKOFFDELAY}, which counts how many slots have still to come before the next transmission. The value of the register is decremented once per idle slot as requested by the DCF procedure, so the idea is to start with an empty bitmap and mark those slots for which the countdown procedure was interrupted.

To avoid spurious detection, instead of continuously checking the state of the channel, we rely on the execution of the {\tt rx\_plcp} handler, which is called each time a valid PLCP is detected. When this happens, we implicitly know that the backoff counter has been frozen at least $20\mu s$ ago, which is the time between the detection of the first short trailling sequence and the complete decoding of the PLCP signal data. In any case, the slot is marked as busy.

We have the option to keep filling the bitmap for up to {\tt BITMAP\_ROUND\_BUILD} consecutive successful transmissions ($\gamma$ in Section~\ref{schedReset}) before checking if the central slot in the bitmap is available (a schedule \emph{halving}), and if that is the case we perform a halving of the node's current schedule.

We performed four experiments for every number of contenders, starting from 1 up to 25. For every experiment each station establishes an iPerf~\cite{tirumala2005iperf} session and transmits saturated UDP traffic towards a central AP, which also served as iPerf server for each flow. We used WiFi channel 14  in order to avoid interference from other networks. The aggregated throughput is derived from the iPerf logs, while the percentage of lost frames is reported by the firmware. This is known to be $(\text{tx}-\text{sx})/\text{tx}$; where tx are the number of transmission attempts, and sx are the number of acknowledged frames.

Figure~\ref{fig:throughputImplementation} shows the average aggregated throughput, while Figure~\ref{fig:lossesImplementation} presents the average percentage of lost frames. Details of the testbed are shown in Table~\ref{tab:testbed}.

CSMA/ECA$_{\text{Hys}}$ has greater throughput due to its ability to avoid collisions more efficiently than CSMA/CA. Further, Schedule Reset (or halving in this case) prevents CSMA/ECA$_{\text{Hys}}$ nodes from increasing too much the time between successful transmissions. Nevertheless, the implementation results reveal that there are other underlying factors that disrupt collision-free schedules. This is evidenced by the increasing percentage of lost frames followed by a throughput degradation in CSMA/ECA$_{\text{Hys}}$. The analysis of the factors that may disrupt collision-free schedules on real hardware implementations of CSMA/ECA$_{\text{Hys}}$ is left as a future research topic.

	\begin{table}
		\centering
		\caption{PHY, MAC and other parameters for the real hardwre experiments \textcolor{red}{REVIEW}}
		\label{tab:testbed}
		\begin{tabular}{|c|c|}
			\hline
			\multicolumn{2}{|c|}{{\bfseries PHY}}\\
			\hline
			{\bfseries Parameter} & {\bfseries Value}\\
			\hline
			PHY rate & 48~Mbps\\
			Empty slot & $9~\mu s$\\
			DIFS & $28~\mu s$\\
			SIFS & $10~\mu s$\\
			IEEE 802.11g WiFi channel & 14\\
			\hline
			\multicolumn{2}{|c|}{{\bfseries MAC}}\\
			\hline
			{\bfseries Parameter} & {\bfseries Value}\\
			\hline
			Maximum backoff stage ($m$) & 5\\
			Minium Contention Window ($CW_{\min}$) & 16\\
			Maximum retransmission attempts & 6\\
			Data payload (Bytes) & 1470\\
			Schedule Reset $\gamma$ & 1\\
			Schedule Reset mode & halving, dynStick\\
			\hline
			\multicolumn{2}{|c|}{{\bfseries TESTBED}}\\
			\hline
			N & 25\\
			Distance between nodes and AP & 8~m\\
			Arrangement of nodes & Semicircle\\
			\hline
		\end{tabular}
	\end{table}

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/throughput-sat-SR-IMPLEMENTATION.eps}
		\caption{Average aggregated throughput for a saturated network. Real hardware implementation results (see Table~\ref{tab:testbed})}
		\label{fig:throughputImplementation}
	\end{figure}

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/losses-sat-SR-IMPLEMENTATION.eps}
		\caption{Average percentage of losses for a saturated network. Real hardware implementation results (see Table~\ref{tab:testbed})}
		\label{fig:lossesImplementation}
	\end{figure}

>>>>

We confirmed the simulations results experimentally with a testbed of real prototypes. This phase was mandatory in order to check whether the behaviour of the system is still correct in the presence of unexpected phenomena, like excessive channel errors, or technical limitations of the hardware like imperfect node synchronisation; and to verify the implementation feasibility of the Schedule Reset mechanism with a real time horizon.

As CSMA/ECA$_{\text{Hys}}$ works with the standard 802.11 PHY, we do not need the flexibility offered by complex development kits based on FPGAs like the WARP boards~\cite{amiri2007warp}. Any architecture that allows a customisation of the channel access delay would fit our requirements. For this reason we chose a widely available 802.11b/g WNIC from Broadcom. Like many other cards, its behaviour is ruled by a firmware code that controls the underlying hardware (including the radio, carrier sense and FIFOs) in real time. In particular we used the BRCM4318KBFG PCI card as it is supported by the OpenFWWF~\cite{OpenFWWF} open-source firmware, an alternative to the original code from the manufacturer that has been recently considered as development platform in several research projects~\cite{WMP,gringolitmc14,gringoliccr14,berger14,CF-MAC}.

Thanks to the extremely low price of this card (refurbished devices are widely available for less than $\$10$ each), it is possible to run experiments with very large testbeds. We used 25 PC-Engines Alix 2d2 nodes running Linux kernel 2.6.32 as stations and a more powerful computer with the same kernel as Access Point (AP), all nodes being equipped with the aforementioned WNIC.

@@

To incorporate stickiness into the prototype we added a {\tt STATE} to the system that can be either {\tt STATE\_DETERMINISTIC} or {\tt STATE\_RANDOM} (related to the type of backoff being used), and a {\tt STICKINESS} counter that we reset each time we have a success and decrement when failing. When the counter gets to zero we enter the random state. Upon a successful transmission we unconditionally enter the deterministic state and reset the stickiness counter to {\tt DEFAULT\_STICKINESS}.
%Setting this last parameter to zero lets the system operate without the stickiness feature.

For the Schedule Reset mechanism, we added a bitmap for monitoring the state of every single slot after a successful transmission. To index the current slot in the bitmap we used the hardware register {\tt SPR\_IFS\_BKOFFDELAY}, which counts how many slots have still to come before the next transmission. The value of the register is decremented once per idle slot, so the idea is to start with an empty bitmap and mark those slots for which the countdown procedure was interrupted.

To avoid spurious detection, instead of continuously checking the state of the channel, we rely on the execution of the {\tt rx\_plcp} handler, which is called each time a valid PLCP is detected. When this happens, we implicitly know that the backoff counter has been frozen at least $20\mu s$ ago, which is the time between the detection of the first short trailing sequence and the complete decoding of the PLCP signal data. In any case, the slot is marked as busy.

We have the option to keep filling the bitmap for up to {\tt BITMAP\_ROUND\_BUILD} consecutive successful transmissions ($\gamma$ in Section~\ref{schedReset}) before checking if the central slot in the bitmap is available. If that is the case, the node's current schedule is halved (a schedule \emph{halving}, following Algorithm~\ref{alg:schedRest}) and its stickiness is incremented to {\tt DEFAULT\_STICKINESS}$+1$.

We performed four experiments for every number of contenders, starting from 1 up to 25. For every experiment each station establishes an iPerf~\cite{tirumala2005iperf} session and transmits saturated UDP traffic towards a central AP, which also functions as iPerf server for each flow. We used WiFi channel 14  in order to avoid interference from other networks. The aggregated throughput is derived from the iPerf logs, while the percentage of lost frames is reported by the firmware. This is known to be $(\text{tx}-\text{sx})/\text{tx}$; where tx are the number of transmission attempts, and sx are the number of acknowledged frames.

Figure~\ref{fig:implementationResults}a shows the average aggregated throughput, while Figure~\ref{fig:implementationResults}b presents the average percentage of lost frames. Details of the testbed are shown in Table~\ref{tab:testbed}.

CSMA/ECA$_{\text{Hys}}$ has greater throughput due to its ability to avoid collisions more efficiently than CSMA/CA. Further, the Schedule Reset aggressiveness (see Section~\ref{aggr}) prevents CSMA/ECA$_{\text{Hys}}$ nodes from increasing too much the time between successful transmissions.

Nevertheless, the implementation results reveal that there are other underlying factors that disrupt collision-free schedules. This is evidenced by the increasing percentage of lost frames followed by a throughput degradation in CSMA/ECA$_{\text{Hys}}$. The analysis of the factors that may disrupt collision-free schedules on real hardware implementations of CSMA/ECA$_{\text{Hys}}$ is left as a future research topic.

	\begin{table}
		\centering
		\caption{PHY, MAC and other parameters for the real hardwre experiments}
		\label{tab:testbed}
		\begin{tabular}{|c|c|}
			\hline
			\multicolumn{2}{|c|}{{\bfseries PHY}}\\
			\hline
			{\bfseries Parameter} & {\bfseries Value}\\
			\hline
			PHY rate & 48~Mbps\\
			Empty slot & $9~\mu s$\\
			DIFS & $28~\mu s$\\
			SIFS & $10~\mu s$\\
			IEEE 802.11g WiFi channel & 14\\
			\hline
			\multicolumn{2}{|c|}{{\bfseries MAC}}\\
			\hline
			{\bfseries Parameter} & {\bfseries Value}\\
			\hline
			Maximum backoff stage ($m$) & 5\\
			Minium Contention Window ($CW_{\min}$) & 16\\
			Maximum retransmission attempts & 6\\
			Data payload (Bytes) & 1470\\
			Schedule Reset $\gamma$ & 1\\
			Schedule Reset mode & halving, dynStick\\
			Default stickiness & 1\\

			\hline
			\multicolumn{2}{|c|}{{\bfseries TESTBED}}\\
			\hline
			N & 25\\
			Distance between nodes and AP & 8~m\\
			Arrangement of nodes & Semicircle\\
			\hline
		\end{tabular}
	\end{table}

	\begin{figure}[tb]
		\centering
		\includegraphics[width=\linewidth,angle=-90]{figures/tonFigs/implementation-combined.eps}
		\caption{Implementation results: a) Average aggregated throughput for a saturated network. Real hardware implementation results (see Table~\ref{tab:testbed}); b) Average percentage of losses for a saturated network. Real hardware implementation results}
		\label{fig:implementationResults}
	\end{figure}

%	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/throughput-sat-SR-IMPLEMENTATION.eps}
%		\caption{Average aggregated throughput for a saturated network. Real hardware implementation results (see Table~\ref{tab:testbed})}
%		\label{fig:throughputImplementation}
%	\end{figure}
%	
%	\begin{figure}[tb]
%		\centering
%		\includegraphics[width=0.7\linewidth,angle=-90]{figures/tonFigs/losses-sat-SR-IMPLEMENTATION.eps}
%		\caption{Average percentage of losses for a saturated network. Real hardware implementation results (see Table~\ref{tab:testbed})}
%		\label{fig:lossesImplementation}
%	\end{figure}
